\documentclass[preprint]{elsarticle}
% vi:spell spelllang=en

\usepackage{hyperref}

\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}

\usepackage{amsmath}  % Maths
\usepackage{amsfonts} % Maths
\usepackage{amssymb}  % Maths
\usepackage{stmaryrd} % Maths (crochets doubles)

\usepackage{url}     % Mise en forme + liens pour URLs
\usepackage{array}   % Tableaux évolués

\usepackage{comment}


\usepackage{prettyref}
\newrefformat{def}{Def.~\ref{#1}}
\newrefformat{fig}{Fig.~\ref{#1}}
\newrefformat{pro}{Property~\ref{#1}}
\newrefformat{pps}{Proposition~\ref{#1}}
\newrefformat{lem}{Lemma~\ref{#1}}
\newrefformat{thm}{Theorem~\ref{#1}}
\newrefformat{sec}{Sect.~\ref{#1}}
\newrefformat{ssec}{Subsect.~\ref{#1}}
\newrefformat{sssec}{Subsect.~\ref{#1}}
\newrefformat{suppl}{Appendix~\ref{#1}}
\newrefformat{eq}{Eq.~\eqref{#1}}
\newrefformat{ex}{Example~\ref{#1}}
\newrefformat{tb}{Table~\ref{#1}}
\newrefformat{rem}{Remark~\ref{#1}}
\def\pref{\prettyref}

\newdefinition{definition}{Definition}
\newdefinition{property}{Property}
\newdefinition{remark}{Remark}
\newdefinition{proposition}{Proposition}
\newdefinition{example}{Example}

\usepackage{tikz}
\newdimen\pgfex
\newdimen\pgfem
\usetikzlibrary{arrows,shapes,shadows,scopes}
\usetikzlibrary{positioning}
\usetikzlibrary{matrix}
\usetikzlibrary{decorations.text}
\usetikzlibrary{decorations.pathmorphing}

\input{macros}
\input{macros-ph}
\input{macros-grn}

% Commandes À FAIRE
\usepackage{color} % Couleurs du texte
\newcommand{\todo}[1]{\textcolor{red}{\textbf{[TODO: #1]}}}
%\newcommand{\rewrite}[1]{\rewriteil{REWRITE: #1}}


\def\modMF#1{\textcolor{teal}{#1}}
\def\modLP#1{\textcolor{magenta}{#1}}
\def\modMM#1{\textcolor{blue}{#1}}
\def\modOR#1{\textcolor{olive}{#1}}



% Id est
\newcommand{\ie}{i.e.~}

% Césures
\hyphenation{pa-ra-me-tri-za-tion}
\hyphenation{pa-ra-me-tri-za-tions}



\begin{document}

\begin{frontmatter}

\title{\modLP{Identification of} Biological Regulatory Networks from Process Hitting models}
%\\\rewriteil{Or possibly: “Complementarity between PH, IG and Thomas modeling”}

\author[irccyn,nii]{Maxime Folschette}
\ead{Maxime.Folschette@irccyn.ec-nantes.fr}
\author[lri]{Lo\"ic Paulev\'e}
\author[nii]{Katsumi Inoue}
\author[irccyn,nii]{Morgan Magnin}
\author[irccyn]{Olivier Roux}

\address[irccyn]{LUNAM Universit\'e, \'Ecole Centrale de Nantes, IRCCyN UMR CNRS 6597\\
(Institut de Recherche en Communications et Cybern\'etique de Nantes)\\
1 rue de la No\"e - B.P. 92101 - 44321 Nantes Cedex 3, France.}
\address[nii]{National Institute of Informatics,\\
2-1-2, Hitotsubashi, Chiyoda-ku, Tokyo 101-8430, Japan.}
\address[lri]{CNRS, Laboratoire de Recherche en Informatique UMR CNRS 8623\\
Universit\'e Paris-Sud, 91405 Orsay Cedex, France}


\begin{abstract}
\modOR{Qualitative formalisms offer a well-established alternative to the more traditionally used differential equation models of Biological Regulatory Networks (BRNs). These formalisms, introduced by Kauffman, Glas and Thomas, led to numerous theoretical works and practical tools to understand emerging behaviors.} \modMM{The analysis of dynamics of very large models is however a rather hard problem, which led us to previously introduce a new framework:} the Process Hitting (PH), \modMM{which} \modLP{is a particular class of Indeterministic Asynchronous Automata
Network (or safe Petri nets).} \modMM{Its major advantage lies in the efficiency of several static analyses recently designed to assess dynamical properties, making it possible to tackle very large models.}

\modLP{In this paper, we address the formal identification of qualitative
models of Biological Regulatory Networks (BRNs) from PH models.}
First, the inference of the Interaction Graph (IG) from a PH model
\modLP{summarizes the signed influences between the components} that are effective for the dynamics.
Second, we provide the inference of all Ren\'e Thomas models of BRNs 
that are compatible with a given PH.
As the PH allows the specification of indeterministic interactions between
components, our inference emphasizes the ability of PH to deal with large BRNs with incomplete knowledge
on interactions, where Thomas' approach fails because of the combinatorics of parameters.

The inference of corresponding Thomas models is \modLP{implemented} using Answer Set Programming,
which allows an efficient enumeration of (possibly numerous) compatible parametrizations.
\end{abstract}

\begin{keyword}
qualitative modeling \sep
model abstraction \sep
interaction graph \sep
René Thomas' parameters \sep
answer set programming
\end{keyword}

\end{frontmatter}


\input{parts-intro}

\section{Frameworks}\label{sec:frameworks}
\input{parts-ph}
\input{parts-grn}

\input{parts-tr2global}
\input{parts-infer-ig}
\input{parts-infer-param}
\input{parts-asp}
\input{parts-example}
\input{parts-discussion}

\bibliographystyle{elsarticle-num}
\bibliography{biblio}

\appendix
\input{annex-phage-lambda}

\end{document}
